### 2021-04-25 paper/parlr_paper_1.tex
faster SimulateComparisons function

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Most methods for conducting record linkage, the task of identifying
duplicate records across datasets, are derived from the foundational
article by Fellegi and Sunter in 1969. Within this family of methods,
two methods have notable strengths: \texttt{fastlink}, proposed by
Enamorado et al (2019), continues a is able to conduct linkage on
incredibly large datasets

@@

Our notation and assumptions closely follow that of Sadinle (2017).
Denote two files as \(A\) and \(B\), with \(n_A\) and \(n_B\) records
respectively, and with records indexed as \(i \in \{1, \ldots, n_A\}\)
in \(A\) and \(j \in \{1, \ldots, n_B\}\) in \(B\). Without loss of
generality, label the files such that \(n_A \geq n_B\). We also assume
there are no duplicates within files, only across. For each record pair
under consideration, we generate a comparison vector
\(\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}\),
where \(F\) is the number of fields used in the linkage and each
\(\gamma_{ij}^f\) takes on a value \(l \in \{1, \ldots, L_f\}\)
indicating level agreement between the two records on a specified field.

@@

@@

Broadly speaking, we increase our computational efficiency by
recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. Let \(H\)
be the set of unique agreement patterns in the data, let \(P\) denote
the total number of unique agreement patterns. Note that \(P\) is
bounded above by \(\prod_{f=1}^F L_f\), and that this bound does not
scale with \(n_A\) or \(n_B\). We index these agreement patterns by
\(p \in \{1, \ldots, P\}\), and say \((i,j) \in h_p\) when the \((i,j)\)
pair exhibits the \(p^{th}\) agreement pattern. Wherever possible, we
conduct calculations over these \(P\) agreement patterns rather than the
\(n_A \times n_B\) record pairs.

\hypertarget{data-representation-and-hashing}{%
\subsection{Data Representation and
Hashing}\label{data-representation-and-hashing}}

In the classic Fellegi Sunter framework, \(\Gamma\) is a
\(n_A n_B \times F\) matrix, with each row providing the comparison
vector for a different \((i,j)\) pair. We however do not store these
comparison vectors themselves, but instead only the a hashed value
corresponding to the agreement pattern of the \((i, j)\) pair. Enamorado
et al (2019) provided the hashing function
\[\sum_{f=1}^F \mathbf{1}_{\gamma_{(i,j)}^f >0}2^{\gamma_{(i,j)}^f + \mathbf{1}_{f>1} \times \sum_{e=1}^{f-1}(L_e-1)}\]
to map each agreement pattern to a unique value, but packages like
\texttt{dplyr} in \texttt{R} are capable of this as well.

We store this information in a nested list \(\tilde{\Gamma}\) where the
\(p^{th}\) component of the \(j^{th}\) list contains a vector of records
in \(A\) that share agreement pattern \(h_p\) with record \(j \in B\).
For each \(p\), we also calculate
\(|h_p| = \sum_{i=1}^{n_A}\sum_{j=1}^{n_B} \mathbf{1}_{(i,j) \in h_p}\)
total instances of agreement pattern \(h_p\) throughout the data, and
also for each \(j\), we calculate
\(|h_p|_j = \sum_{i=1}^{n_A} \mathbf{1}_{(i,j) \in h_p}\) the instances
of agreement pattern \(p\) among the comparison vectors between record
\(j \in B\) and each of the \(n_A\) records in \(A\).

For large data, we can partition the two datasets \(A\) and \(B\) into
smaller blocks \(\{A_m\}\) and \$\{B\_m\} for more manageable
computations. On a single machine, we can read-in data sequentially,
conduct hashing, collect results, and delete the original data from
memory before continuing with the next chunk of data. With multiple
cores or multiple machines, this can be done in parallel. Storing this
hashed information still becomes burdensome for large data, but this
hashing method greatly expands the capabilities of the Fellegi-Sunter
framework.

Lastly, the classic Fellegi Sunter method represents the \(\gamma_{ij}\)
comparison vector as vector of length \(F\), with each component
\(\gamma_{ij}^f\) taking on values in \(\{0, \ldots, L_f - 1\}\). To
ease computations, we instead represent the comparison as a
concatenation of \(F\) many binary indicator vectors of lengths \(L_f\).
For example, if \(L_1 = L_2 = 2\) and \(L_3 = 3\), then
\(\gamma_{ij} = (1, 0, 2)\) under the classical framework becomes
\(\gamma_{ij} = (0, 1, 1, 0, 0, 0, 1)\) under our framework. This is a
bijective transformation that does not change the meaning of the data,
but this representation will ease calculations and posterior updates.

@@

\begin{itemize}
\item
  Replicating the Sadinle Study
\item
  \texttt{parlr} and \texttt{BRL} speed comparison
\end{itemize}

>>>>

\hypertarget{related-work}{%
\subsection{Related Work}\label{related-work}}

Most record linkage techniques are derived from the seminal 1969 paper
by Fellegi and Sunter: ``A Theory for Record Linkage''. The defining
characteristic of their model was to transform the sets of records,
which often contain text data that is difficult to model, into sets of
comparison vectors governed by parameters that can be more easily
estimated. Concretely, if files \(A\) and \(B\) have \(n_A\) and \(n_B\)
records respectively, and if the files share \(F\) fields in common upon
which to base the linkage, the Fellegi and Sunter approach generates a
\(n_A n_B \times F\) matrix \(\Gamma\), which contains similarity scores
between each pair of records across datasets. We say \(\gamma_{ij}\) is
the comparison vector for record \(i \in A\) and record \(j \in B\),
with \(\gamma_{ij}^f\) providing their similarity score on the
\(f^{th}\) field. For ease of modeling and computation, we restrict
these similarity scores to be discrete, ordinal variables, and the
construction of these is left to the modeller. It is common to use use
binary 0-1 variables to indicate exact matching, and 0-1-2 variables to
provide an option for partial matching. For text data, we calculate
similarity based on Levenstein distance or some other text similarity
score, and bin these scores to integers for use in the model.

Two modern adaptations of the Fellegi Sunter model are important for
understanding \texttt{parlr} and its contribution. In 2019, Enamorado et
al proposed \texttt{fastlink}, a method that closely followed the
modelling assumptions of Fellegi and Sunter, but used innovating hashing
techniques to greatly enhance its speed and scale. Importantly, they
maintained Fellegi and Sunter's independent matching assumption, in
which the matching status of any \((i,j)\) pair is made independently of
the matching of any other \((i', j')\). This leads to many matchings
that violate ``one-to-one'' considerations that are often implicit in
the data, which need to be resolved in a post-processing step. In
contrast, Sadinle (2017) proposed ``Beta Record Linkage for Bipartite
Matching'' and the accompanying \texttt{BRL} package which strictly
enforces one-to-one matching. Specifically, in each iteration of his
Gibbs sampler, he considers each record \(j\in B\), removes from
consideration the records \(i\in A\) that have already been matched,
then samples a potential link. However, accounting for these
dependencies throughout the linkage process is computionally burdensome,
leaving \texttt{BRL} only suitable for small to moderate linkage
problems.

Outside of the Fellegi-Sunter framework, some researchers have developed
methods that model the data directly, not a derived set of comparison
vectors. As an early example, Steorts et al 2016 produced
\texttt{blink}, which had the advantage of a computational complexity
that grew linearly with the data (as opposed to quadratically within
Fellegi Sunter), but was plagued with slow mixing times and was unable
to incorportate text data. Later, Steorts et al 2020 improved on this
method with \texttt{dblink}, which was able to simulate text data within
its Gibbs sampler by drawing from an empirical distribution, and
remedied its slow mixing through a probabalistic blocking approach.
However, this method remains computationally intensive and requires
distributed computing to manage even moderate linkage problems in
acceptable time.

@@

Our notation and assumptions closely follow that of Sadinle (2017).
Denote two files as \(A\) and \(B\), with \(n_A\) and \(n_B\) records
respectively, and with records indexed as \(i \in \{1, \ldots, n_A\}\)
in \(A\) and \(j \in \{1, \ldots, n_B\}\) in \(B\). Without loss of
generality, label the files such that \(n_A \geq n_B\). We also assume
there are no duplicates within files, only across. For each record pair
under consideration, we generate a comparison vector
\(\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}\),
where \(F\) is the number of fields used in the linkage and each
\(\gamma_{ij}^f\) takes on a value \(l \in \{1, \ldots, L_f\}\)
indicating the level agreement between the two records on a specified
field.

@@

Since our Gibbs procedure does not strictly enforce one-to-one matching,
it is possible for the final Bayes estimate to link mulitple records in
\(B\) to one record in \(A\). The modeller can either report both such
matches (with their respective posterior match probabilities), or
resolve these conflicts by accepting only the match with highest
posterior probability. Rigorous proof of this resolution procedure is
deferred to future work.

@@

Broadly speaking, we increase our computational efficiency by
recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. Let
\(\mathcal{H}\) be the set of unique agreement patterns in the data, let
\(P\) denote the total number of unique agreement patterns. Note that
\(P\) is bounded above by \(\prod_{f=1}^F L_f\), and that this bound
does not scale with \(n_A\) or \(n_B\). Prior to processing the data, we
identify all \(P\) patterns in \(\mathcal{H}\) and enumerate them
\(h_1, \ldots, h_P\). When the \((i,j)\) pair exhibits the \(p^{th}\)
agreement pattern, we say \((i,j) \in h_p\). Wherever possible, we
conduct calculations over these \(P\) agreement patterns rather than the
\(n_A \times n_B\) record pairs.

\hypertarget{data-representation-hashing-and-storage}{%
\subsection{Data Representation, Hashing, and
Storage}\label{data-representation-hashing-and-storage}}

The classic Fellegi Sunter method represents the \(\gamma_{ij}\)
comparison vector as vector of length \(F\), with each component
\(\gamma_{ij}^f\) taking on values in \(\{0, \ldots, L_f - 1\}\). To
ease computations, we instead represent the comparison as a
concatenation of \(F\) many binary indicator vectors of lengths \(L_f\).
For example, if \(L_1 = L_2 = 2\) and \(L_3 = 3\), then
\(\gamma_{ij} = (1, 0, 2)\) under the classical framework becomes
\(\gamma_{ij} = (0, 1, 1, 0, 0, 0, 1)\) under our framework. This is a
bijective transformation that does not change the meaning of the data,
but this representation will ease calculations and posterior updates.

In the classic Fellegi Sunter framework, \(\Gamma\) is a
\(n_A n_B \times F\) matrix, with each row providing the comparison
vector for a different \((i,j)\) pair. We however do not store these
comparison vectors themselves, but instead only a hashed value
corresponding to the agreement pattern of the \((i, j)\) pair. We store
this information in a nested list \(\tilde{\Gamma}\) where the
\(p^{th}\) component of the \(j^{th}\) list contains a vector of records
in \(A\) that share agreement pattern \(p\) with record \(j \in B\). For
each \(p\), we also calculate
\(H_p = \sum_{i=1}^{n_A}\sum_{j=1}^{n_B} \mathbf{1}_{(i,j) \in h_p}\),
the total instances of agreement pattern \(p\) throughout the data, and
also for each \(j\), we calculate
\(H_{p_j} = \sum_{i=1}^{n_A} \mathbf{1}_{{(i,j) \in h_p}}\) the
instances of agreement pattern \(p\) among the comparison vectors
between record \(j \in B\) and each of the \(n_A\) records in \(A\).

For large data, we can partition the two datasets \(A\) and \(B\) into
smaller blocks \(\{A_m\}\) and \(\{B_m\}\) for more manageable
computations. On a single machine, we can read-in data sequentially,
conduct hashing, collect results, and delete the original data from
memory before continuing with the next chunk of data. With multiple
cores or multiple machines, this can be done in parallel. Storing this
hashed information still becomes burdensome for large data, but this
hashing method greatly expands the capabilities of the Fellegi-Sunter
framework.

The hashing procedure described above considerably reduces the memory
needed to store the comparison information: instead of storing
\(n_A \times n_B\) comparison vectors, which are relatively long under
either the Fellegi Sunter or our modified framework, we only store the
\(P\) unique vectors, and then \(n_A \times n_B\) scalar quantities
relating record pairs to those vectors. However, even storing these
\(n_A \times n_B\) scalar labels can become burdensome with large data.
Worse, the overwhelming majority of these labels relate to record pairs
that are clear non-matches.

To address this, we propose a new strategy called \emph{random
indexing}. In standard indexing, the modeller decides a certain criteria
that they expect all true matching pairs to satisfy, and a priori label
any record pairs that do not meet that criteria as non-matching. For
example, one might only consider pairs with a certain similarity score
on a field deemed to be important (like first name), or only pairs with
exact matching on a specified number of fields. While generally chosen
to be be quite loose, establishing these criteria requires knowledge of
the problem and invites room for human error. We propose a method of
reducing the comparison space (and reducing storage requirements)
without these drawbacks. Note that all records of the same agreement
pattern have the same probality when sampling \(Z_j\). Therefore we know
that records belonging to an \(h_p\) such that \(H_{p_j}\) is large are
very unlikely to be sampled consistently enough to be deemed a match
through the Bayes estimate, even without considering the form of the
agreement pattern itself. In random indexing, rather than store all of
these unlikely record labels, we choose to store only a small number
\(R\) of them. Posterior calculations still attribute the appropiate
weight to all records through the summary statistics \(H_{p_j}, H_p^m\),
and \(H_p^u\); this is similar to the \emph{u-correction} concept
introduced by \_\_\_\_\_\_\_\_. Rather than storing \(n_A \times n_B\)
record labels, random indexing allows us to store at most
\(n_B \times P \times R\), regardless of how large \(n_A\) is. Coupled
with the partitioning strategy for creating comparison vectors, this
greatly expands the size of the linkage tasks we can undertake.

@@

In his 2017 paper, Sadinle demonstrated the strength of his method by
running \texttt{BRL} on simulated pairs of files with differing levels
of errors and overlap. Comparing our method against \texttt{BRL} on
these same simulated datasets, we find that our method only has weakened
performance in the extreme scenario of very high errors and very high
overlap across files. We use the following measures of accuracy in our
analysis: although using only recall and precision is common in the
literature, we find it useful to have a single metric the gauge overall
performance, and thus also include F measure.

\[\text{Recall} = \frac{\text{Matches Correctly Identified}}{\text{True Matches in Data}}\]
\[\text{Precision} = \frac{\text{Declared Matches}}{\text{True Matches in Data}}\]
\[\text{F-Measure} = 2\left(\frac{\text{Recall} \times \text{Precision}}{\text{Recall} + \text{Precision}}\right)\]

\begin{center}\includegraphics[width=29.17in]{../notes/figures/sadinle_sim_plot} \end{center}

To demonstrate speed, we generate comparison vectors from pre-specified
distributions so that we can easily increase the size of the linkage
problem. Distributions are meant to emulate the behavior of similarity
scores across first name, last name, and day, month, and and year of
birth, with 1 indicating an exact match on a particular field and 0
indicating a nonmatch. We simulate these data for different values of
\(n_A\) and \(n_B\), and compare the runtime of \texttt{parlr} against
\texttt{BRL}. We see that at low data size, \texttt{BRL} outperforms,
but that \texttt{parlr} is significantly faster at handling larger data.

In particular, runtime for \texttt{BRL} seems to grow quadratically
while runtime for \texttt{parlr} seems to grow linearly.

\begin{table}[ht]
\centering
\begin{tabular}{rrr}
  \hline
 & m & u \\ 
  \hline
fname\_0 & 0.05 & 0.99 \\ 
  fname\_1 & 0.95 & 0.01 \\ 
  lname\_0 & 0.05 & 0.99 \\ 
  lname\_1 & 0.95 & 0.01 \\ 
  day\_0 & 0.05 & 0.97 \\ 
  day\_1 & 0.95 & 0.03 \\ 
  month\_0 & 0.05 & 0.92 \\ 
  month\_1 & 0.95 & 0.08 \\ 
  year\_0 & 0.05 & 0.93 \\ 
  year\_1 & 0.95 & 0.07 \\ 
   \hline
\end{tabular}
\end{table}

\begin{center}\includegraphics[width=23.6in]{../notes/figures/sadinle_speed_plot} \end{center}

The above discussion suggests that for fixed \(n_B\), computation time
should remain mostly constant with growing \(n_A\). Shockingly, this
seems to be true. In the plot below, fixing \(n_B = 500\), we see linear
growth for the runtime under \texttt{BRL} as \(n_A\) increases, with
much more static runtime under \texttt{parlr}. The slight increases in
runtime that we do see are due primarily to the hashing step, which
again can be run in parallel for large data.

\begin{center}\includegraphics[width=29.17in]{../notes/figures/speed_plot_fixed_nB} \end{center}

We note here that \texttt{BRL} is coded in C, which makes for unfair
comparison against \texttt{parlr}, currently only built in R.
Additionally, although \texttt{parlr} is amenable to parallezation, this
simulation was run on a single core. Running \texttt{parlr} in C++ with
paralellization for the hashing step and sampling the matching status of
the record pairs should lead to even more drastic results.
